# Introduction to Programming with R

## Table of Contents

* [RStudio IDE](#rstudio-ide)
* [Installation of R Packages](#installation-of-r-packages)
* [R Datastructures](#r-datastructures)
* [Data Manipulation](#data-manipulation)
* [Descriptive Statistics](#descriptive-statistics)
* [Graphical Exploration](#graphical-exploration)
* [String Processing with R](#strings)
* [Mathematical operations](#math)
* [User-defined functions](#user-defined-functions)
* [Flow control structures](#flow-control)
* [Apply-family functions](#apply-family-functions)
* [Next steps](#further-readings)

## RStudio IDE
[RStudio](https://www.rstudio.com/) is an advanced IDE for the R programming language, web development with [Shiny](http://shiny.rstudio.com/),
literate scientific programming and authoring rich data presentations.

Let's look at this masterpiece!

### Getting help
Two very first useful commands are `?` (search) and `??` (fuzzy search).

```{r, eval = FALSE}
?plot
??ggplot2
help(plot)
```

### Current working directory

```{r, eval = FALSE}
getwd()

# setwd('some/path')
```

You can use the `Session` menu to manipulate the path to the working directory.

### Closing notes
`RStudio` isn't the only way to work within an R environment (also it's admittedly a very convenient one).
Other options could be:

- plain R console;
- RCommander (the package `rcmdr`);
- ESS (for Emacs users);
- Tinn-R.

This list isn't an exhaustive one, look please at http://www.sciviews.org/_rgui/.

## Installation of R Packages

During both sessions we'll need some additional libraries. R is a very powerful language, but the real
power comes from reusing ready code from so called "packages".

Required packages are:

- `ggplot2` -- a modern plotting infrastructure;
- `tm` -- term matrix;
- `Matrix` -- sparse matrix representation;
- `topicmodels` -- LDA models;
- `wordcloud` -- graphical representation of word-clouds;
- `gtools` -- R addendum to standard functions;
- `moments` -- useful descriptive statisctics;
- `car` -- companion to applied regression;
- `caret` -- classification and regression training;
- `glmnet` -- generalized linear models.

```{r}
libraries <- c('ggplot2', 'tm', 'Matrix', 'topicmodels', 'wordcloud', 'gtools', 'moments', 'car', 'caret', 'glmnet')

missing_libraries <- libraries[!(libraries %in% installed.packages()[, 'Package'])]

if (length(missing_libraries) > 0) {
  install.packages(missing_libraries, repos = 'http://cran.rstudio.com/')
}
```

Before we can use a library, we have to express our wish to do so:

```{r, eval = FALSE}
library(ggplot2)
```

This should be done only once for each required package.

### Closing notes
We can use the `Packages` menu for installing additional libraries.

A modern approach to manage all project dependencies is a package manager. In the R ecosystem it is
[Packrat](http://rstudio.github.io/packrat/).

## <a name="r-datastructures"></a>Basic datastructures in R
`R` has a specialization in processing collections of values. That's why we cannot find any `scalar`
built in data structure. Every data structure is considered a collection even if it contains only
one value.

### Vectors
A primary data structure is a vector: a collection of values of the `same` type.

Which types are built in:

- integer
- double
- logical
- character
- list
- etc.

```{r, eval = FALSE}
v1 <- c(18.3, 19.4, 18.5, 17.2, 22.5, 17.8, 23.4)
typeof(v1)
is.vector(v1)
```

We can access different elements of our vector:
```{r, eval = FALSE}
v1[1]
v1[-3]
v1[c(2, 3, 5)]
v1[c(T, T, T, F, F, T, T)]
```

We can compare different vectors:
```{r, eval = FALSE}
1 < 2

v2 <- seq(14, 20, by = 2)

v3 <- 14:17

v2 < v3
```

### Lists
Lists are in R data-structures for storing heterogeneous data of (possibly) different length.

```{r, eval = FALSE}
l1 <- as.list(v1)
l2 <- list(c(1, 2, 3, 4), c('a', 'b', 'c', 'd'))
is.list(l1)
is.list(l2)
l1; l2
```

Accessing lists is different that that of verctors:
```{r, eval = FALSE}
l1[[1]]
l2[[2]][1]
```

### Matrices
Matrix is two dimensional data structure for objects of the same type.

```{r, eval = FALSE}
m1 <- matrix(1:16, nrow = 4, ncol = 4)
m1[3, 3]
m1[3, ]
m1[, 3]
m1[2:3, 2:3]

is.matrix(m1)
```

Matrices are very efficient in storing numerical data.

### Data Frames
Data Frames are very useful as a metaphor for data tables (like Excel datasheets).

Mostly we become new dataframes while reading in data from external sources.
But we can create dataframes ourselves:
```{r, eval = FALSE}
d1 <- data.frame(small_values = v2, bigger_values = v3)
d1['biggest_values'] <- 45:48
d1['letters'] <- c('a', 'b', 'c', 'd')
d1[2, 1] <- 0.3
d1$letters[2]
d1[2, ]
d1
```

Let's create some useful data frame:


    POS       TokenFrequency  TypeFrequency	Class
    ADJ	      421		          271		        open
    ADV       337		          103		        open
    N	        1411	          735		        open
    KONJ      458		          18		        closed
    PRAEP     455		          37		        closed


```{r, eval = FALSE}
pos <- c('ADJ', 'ADV', 'N', 'KONJ', 'PRAEP')

token_freq <- c(421, 337, 1311, 458, 455)
type_freq <- c(271, 103,735,  18, 37)
class <- c('open', 'open', 'open', 'closed', 'closed')

d2 <- data.frame(POS = pos, TokenFrequency = token_freq, TypeFrequency = type_freq, Class = class)

```

### Special values

There are two very insteresting data types in R: `NULL` and `NA`.

```{r, eval = FALSE}
c(1, 2.1, T, NA, 4)
c(1, 2, T, NA, 4)
```


### Closing notes
`R` is an object oriented language with functional elements, so we can imagine plenty of different
data-structures.

Not mentioned here are `arrays`, `factors`, different data types etc.

## Data manipulation
The real power of `R` unfolds with the manipulation of tabular data.

### Reading data
We'll try to read in a data set from a `.csv` file:
```{r, cache = TRUE}
pause_times <- read.table('data/pauses.txt', header = T)
head(pause_times)
```

### Writing data
To write a dataframe into a file use:
```{r, eval = FALSE}
write.csv2(d2, file = 'data/d2.csv')
write.table(d2, file = 'd2.tsv', sep = '\t')
```

### Subsetting data

```{r samples, echo = TRUE, eval = FALSE}
set.seed(1)
d2[sample(length(d2$Class)), ]
d2[sample(length(d2$Class)), ]
```

### Boolean Queries

```{r boolean, echo = TRUE, eval = FALSE}
subset(d2, d2$Klasse == 'open'   & d2$TokenFrequency > 100)
subset(d2, d2$Klasse == 'closed' & d2$POS %in% c('KONJ', 'PRAEP'))
```

### Ordering and sorting

```{r sort, echo = TRUE, eval = FALSE}
v12 <- 1:100 # seq(1, 100, 1)
v13 <- c(1, 4, 5, 3, 6, 9)
sort(v13)
sort(v13, TRUE)
order(v13)
```

```{r ordering, echo = TRUE, eval = FALSE}
d2[order(d2$TokenFrequency), ]
```

### Closing notes
A convenient way to load standard datasets is the menu `Import Dataset`.

You may be interested in SQL connectivity since many data sets are huge. Consider the package `sqldf`.

Very useful are the modern packages for data manipulation like `dplyr`, `tidyr`.

## Desctiptive Statistics

```{r descriptive_measures, eval = FALSE}
library(moments)

random.vector <- rnorm(100)
str(random.vector)
summary(random.vector)
mean(random.vector)
median(random.vector)
quantile(random.vector, c(0.25, 0.5, 0.75))
sd(random.vector)
var(random.vector)
mad(random.vector)

skewness(random.vector)
kurtosis(random.vector)
```

### Closing notes
You may be interested in SQL connectivity since many data sets are huge. Consider the package `sqldf`.

## Graphical exploration 
with base and ggplo2 systems

### Frequency Distributions
```{r table, echo = T, eval = FALSE}
head(pause_times$FILLER)

attach(pause_times)

str(FILLER)

table(FILLER)

table(FILLER) / length(FILLER)

prop.table(table(FILLER))

cumsum(table(FILLER))

detach(pause_times)
```

### Piecharts

```{r piechart, echo = TRUE, eval = FALSE}
pie(table(pause_times$FILLER))
```


### Barcharts

```{r barplot, echo = T, eval = FALSE}
barplot(table(pause_times$FILLER))
barplot(table(pause_times$FILLER),
        main = 'Our Barchart',
        # ylim
        # xlim
        col = c('red', 'green', 'blue'),
        names.arg = c('Aeh', 'Aehm', 'Silence')
        )

```


### Histograms
```{r hist, echo = T, eval = FALSE} 
hist(pause_times$LENGTH, breaks = c(200,350, 800, 1600))
hist(pause_times$LENGTH, freq = T)
lines(density(pause_times$LENGTH))
```

```{r ggplot_hist, cache = T, eval = FALSE}
library(ggplot2)
qplot(pause_times$LENGTH, binwidth = 5)
ggplot(pause_times, aes(x = LENGTH)) + geom_histogram(binwidth = 15, color = 'black', fill = 'white')
ggplot(pause_times, aes(x = LENGTH)) + geom_density()
```

### Closing notes
`R` has a long history of visualizing data and one of the most elaborated library collections
for that. We have watched at the `base` and `ggplot2` plotting systems. Of interest are `lattice` and
`grid` systems, which can be very useful in particular use cases.

If you want to produce interactive graphics consider using the `ggvis` package or webservices
like [Plotly](https://plot.ly/).

## <a name="strings"></a>String Processing with R

```{r, eval = FALSE}
d2$POS <- tolower(d2$POS)
d2$Class <- gsub('open', 'o', d2$Class)
d2$Class <- substr(d2$Class, 1, 1)

d2
```

### Closing notes
If you want steadily preprocess your data and perform ETL tasks look at the `Stringr` package.

## <a name="math"></a>Doing Math in R
`R` hasn't been thougt primarily for mathematical calculations like e.g. `GNU Octave`,
but you can find a lot of useful functions.

### Arithmetics
```{r, eval = FALSE}
a <- 1
b <- 2
a + b
a - b
a * b
a / b
a ^ b
5 %/% 2
5 %% 2
```

### Sums and Products
```{r, eval = FALSE}
x <- 1:10
sum(x)
prod(x)
factorial(length(x))
```

### Roots
```{r, eval = FALSE}
sqrt(4)
8^(1/3)
```

### Log Functions

```{r, eval = FALSE}
pi
exp(1)
log(1, base = exp(1))
log10(100)
log2(8)
```

### Trigonometric Functions

Measuring Angles in Radians:
$$
  \alpha^{\circ} = \frac{\pi}{180}\cdot \alpha \text{ rad} 
$$

```{r, eval = FALSE}
pi
sin((3/2)*pi)
cos(0)
cos(pi/3)
tan(pi/4)
```


### Unions and Intersections
```{r, eval = FALSE}
union(1:10,5:7)
intersect(1:10,5:7)
```

### Matrices
```{r, eval = FALSE}
A <- matrix(1:6, nrow = 2, ncol = 3, byrow = T)
A
t(A)
```

#### Matrix Addition
```{r, eval = FALSE}
B <- matrix(6:11, nrow = 2, ncol = 3, byrow = T)
B
A + B
A - B
```

#### Matrix Multiplication
```{r, eval = FALSE}
A
A * 2

D <- matrix(6:11, nrow = 3, ncol = 2, byrow = T)
A; D

A %*% D
```

## User-defined functions
```{r, eval = FALSE}
seq_up_to <- function(x) {
  seq(1, as.double(x))
}

seq_up_to(13)
```

### Closing notes
Write your own functions whenever you feel comfortable with.

## <a name="flow-control"></a>Flow control structures

```{r, eval = FALSE}
ifelse(d2$Class == 'c', 'Closed Class', 'Open Class')

if (d2$Class == 'c') {
  'Closed Class'
} else {
  'Open Class'
}
```

```{r, eval = FALSE}

for (fruit in c('Apple', 'Banana', 'Melone')) {
  print(tolower(fruit))
}

```


## Apply-family functions

```{r, eval = FALSE}
lapply(list(seq_up_to(100), seq_up_to(1000), seq_up_to(10000)), sum)
```


### Closing notes
Use apply functions everywhere instead of loops since verctorization in R works much faster.

## <a name="further-readings"></a>Where to go from here

The primary source of information should be the official documentation:

- https://www.r-project.org/
- https://cran.r-project.org/web/views/

Some online resources are well grounded and useful:

- https://github.com/qinwf/awesome-R
- http://adv-r.had.co.nz/Style.html

If you were interested in the topics of the current presentation you may want to read ahead.
Some sources will help you to master the R Language and it's ecosystem:

- Horton, Nicholas J., and Ken Kleinman. Using R and RStudio for Data Management, Statistical Analysis, and Graphics. 2nd ed. Boca Raton, FL: CRC Press, Taylor & Francis Group, 2015.
- Matloff, Norman. The Art of R Programming: A Tour of Statistical Software Design. 1st ed. No Starch Press, 2011.
- Verzani, John. Using R for Introductory Statistics. Boca Raton: Chapman & Hall/CRC, 2005. 
- Wickham, Hadley. Advanced R. The R Series. Boca Raton, FL: CRC Press, Taylor & Francis Group, 2015.
- Wickham, Hadley. ggplot2: Elegant Graphics for Data Analysis. Use R! Springer Science   Business Media, 2009.

